``Interestingness'' in Mathematical Theory Exploration

### 2018-05-17 PhDThesis/litreview_mathhistory.tex
Drag a bunch of stuff into thesis folder

% Derived from the template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{csvsimple}
\usepackage{float}
\restylefloat{table}
\usepackage{hhline}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
  shapes.geometric}
% etc.

\usepackage{attrib}
\usepackage{csquotes}

\usepackage{graphicx}
\graphicspath{ {support/} }

% please place your own definitions here and don't use \def but
% \newcommand{}{}

\newcommand{\etal}{{\em et al.}}
\newcommand{\Bagai}{Bagai \etal{}}

% For example theory of lists
\newcommand{\function}{\rightarrow}

\newcommand{\Zero}{\text{Z}}
\newcommand{\Succ}{\text{S}}
\newcommand{\plus}{\text{plus}}
\newcommand{\mult}{\text{times}}

\newcommand{\List}{\text{List}}
\newcommand{\ListA}{\text{List} \  a}
\newcommand{\Nil}{\text{Nil}}
\newcommand{\Cons}{\text{Cons}}
\newcommand{\Head}{\text{head}}
\newcommand{\Tail}{\text{tail}}
\newcommand{\Append}{\text{append}}
\newcommand{\Reverse}{\text{reverse}}
\newcommand{\Length}{\text{length}}
\newcommand{\Map}{\text{map}}
\newcommand{\Foldl}{\text{foldl}}
\newcommand{\Foldr}{\text{foldr}}

% For interestingness table
\newcommand{\iE}{\textbf{E}}
\newcommand{\iN}{\textbf{N}}
\newcommand{\iS}{\textbf{S}}
\newcommand{\iA}{\textbf{A}}
\newcommand{\iC}{\textbf{C}}
\newcommand{\iU}{\textbf{U}}
\newcommand{\tIFF}{if-and-only-if}
\newcommand{\tNE}{non-exists}
\newcommand{\tIMP}{implies}
\newcommand{\tEQ}{equations}
\newcommand{\tINE}{inequalities}
\newcommand{\tCON}{conditional}
\newcommand{\tRow}[1]{#1 \\ \hline}

% Insert the name of "your journal" with
\journalname{Journal of Automated Reasoning}

\begin{document}

\title{Quantitative Benchmarking for Automatically Generated Conjectures%\thanks
% {Grants or other notes about the article that should go on the front page
% should be placed here. General acknowledgments should be placed at the end of
% the article.}
}
% \subtitle{Do you have a subtitle?\\ If so, write it here}

% \titlerunning{Short form of title}        % if too long for running head

\author{Chris Warburton \and
        Alison Pease    \and
        Jianguo Zhang}

% \authorrunning{Short form of author list} % if too long for running head

\institute{C. Warburton \at
           University of Dundee \\
           \email{c.m.warburton@dundee.ac.uk}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           A. Pease \at
           University of Dundee \\
           \email{a.pease@dundee.ac.uk}
           \and
           J. Zhang \at
           University of Dundee \\
           \email{j.n.zhang@dundee.ac.uk}
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

% Look carefully at each tool's (paper's) eval section

% Make it clear that methodology is the point (it's not just a detail, like in a
% chemistry paper for example)

\begin{abstract}
  We propose a benchmarking methodology to evaluate the efficiency and quality
  of \emph{conjecture generation} by automated tools for \emph{mathematical
    theory exploration}. Our approach uses widely available theorem proving
  tasks as a \emph{ground-truth} corpus, and we demonstrate its use on the
  QuickSpec and IsaCoSy tools. We found that both may fail, even for small
  inputs, but QuickSpec usually finishes in significantly less time than IsaCoSy
  and produces significantly more ``interesting'' output. By providing a
  standard, cross-tool evaluation technique we allow the disparate approaches
  currently being pursued to be more directly compared. Our hope is to encourage
  innovation and improvements similar to those seen in fields like automated
  theorem proving, where the availability of benchmarks encourages healthy
  competition.
%\keywords{First keyword \and Second keyword \and More}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}

% Lay out the field of TE: motivation, tools, evaluation, problems...

\emph{Conjecture generation} is the open-ended task of conjecturing properties
of a given logical theory which are somehow ``interesting'', and is
studied as a sub-field of \emph{mathematical theory exploration} (MTE). This has
applications wherever we find uses for formal methods: in proof assistants and
their libraries, in mathematics education and research, and in the specification
and verification of software. Although formal methods, and automated reasoning
more generally, have been advocated in mathematics (notably by
Voevodsky~\cite{voevodsky2010univalent} and Gowers~\cite{ganesalingam2013fully})
and software engineering (for example via dependently-typed
programming~\cite{McKinna:2006}), they have a reputation for being difficult and
expensive, which has historically limited their use to high-assurance domains
such as aerospace and microprocessor design.

One path to reducing these barriers is to increase the level of automation.
Software verification tools have been shown to benefit from the addition of a
conjecture generation step for finding ``background lemmas'', which are useful
when proving user-provided statements~\cite{Claessen.Johansson.Rosen.ea:2013}.
Another path is to tailor the division of labour between humans and machines
such that each is playing to their strengths, which has been referred to as
``centaur teams''~\cite{harari2017reboot,davenport2015beyond}. Machines can
systematically search a much larger space of possible relationships than humans,
potentially bringing to light novel and surprising connections, whilst allowing
the user to make the ultimate judgement over which of the most promising results
are worth investigating further.

Similar tasks are also studied in the domain of Computational \emph{Scientific}
Discovery~\cite{king2004functional,Williams20141289,schmidt2009distilling}, for
example in the search for new drugs. The major difference between scientific and
mathematical discovery is that inductive reasoning and experimental testing are,
in principle, more important for the former, although we note that they are also
core principles in the mathematical tools we have investigated (in particular
for ensuring the \emph{plausibility} of conjectures).

We limit ourselves to mathematical applications, but even here it is difficult
to measure and compare the success of approaches and tools. This is partially
due to their diversity, but also because of the inherent ambiguity of the task
(what counts as ``interesting''?), the different goals emphasised by their
designers and the variety of evaluation methods employed. We attempt to solve
this discrepancy, at least for the foreseeable future, by defining a standard,
unambiguous benchmarking approach with which to compare the conjecture
generation of MTE tools. Our contributions include:

\begin{itemize}
\item A general methodology for benchmarking conjecture generation.
\item Resolving the issue of ``interestingness'' through the use of
  theorem-proving benchmarks as a ground-truth.
\item A specific instantiation of this methodology, using the Tons of Inductive
  Problems (TIP) benchmark as a corpus.
\item Automated tooling to perform this benchmarking.
\item Application of our methodology to the QuickSpec and IsaCoSy MTE tools,
  and a comparison and discussion of the results.
\end{itemize}

We introduce the conjecture generation problem in more detail, along with the
difficulty of comparing existing solutions, in $\S$\ref{sec:background}. We
explain our proposal for a more general benchmark in $\S$\ref{sec:proposal} and
demonstrate its application to existing MTE tools in $\S$\ref{sec:application}.
Issues facing our approach, and the field in general, are discussed in
$\S$\ref{sec:discussion}; related work is surveyed in
$\S$\ref{sec:related-work}, including the diverse terminology found in the
literature; and concluding remarks are given in $\S$\ref{sec:conclusion}.

\begin{sloppypar}
  Our benchmark, including the scripts used to generate it from the TIP problem
  set, is available online at
  \url{https://github.com/warbo/theory-exploration-benchmarks}. The application
  to QuickSpec is at \url{https://github.com/warbo/haskell-te} and IsaCoSy at
  \url{https://github.com/warbo/isaplanner-tip}.
\end{sloppypar}

%- Existing approaches are difficult to compare
% - They want different things
% - They're evaluated differently
% - They're evaluated ``locally'' (as white boxen)

% Look at phd symposium 2017, moa's conjecture synthesis paper, etc.
% We can just jump straight in to automated conjecture generation

\section{Background}
\label{sec:background}

\subsection{Motivation}
\label{sec:motivation}

% more detail about evaluation, interestingness, etc. Reference Simon/Alan
% interestingness, e.g. examples from textbooks, etc.

\begin{figure}
  \begin{equation*}
    \begin{split}
      \forall a. \Nil            &: \ListA                                  \\
      \forall a. \Cons           &: a \rightarrow \ListA \rightarrow \ListA \\
      \Head(\Cons(x, xs))        &= x                                       \\
      \Tail(\Cons(x, xs))        &= xs                                      \\
      \Append(\Nil,         ys)  &= ys                                      \\
      \Append(\Cons(x, xs), ys)  &= \Cons(x, \Append(xs, ys))               \\
      \Reverse(\Nil)             &= \Nil                                    \\
      \Reverse(\Cons(x, xs))     &= \Append(\Reverse(xs), \Cons(x, \Nil))   \\
      \Length(\Nil)              &= \Zero                                   \\
      \Length(\Cons(x, xs))      &= \Succ (\Length(xs))                     \\
      \Map(f, \Nil)              &= \Nil                                    \\
      \Map(f, \Cons(x, xs))      &= \Cons(f(x), \Map(f, xs))                \\
      \Foldl(f, x, \Nil)         &= x                                       \\
      \Foldl(f, x, \Cons(y, ys)) &= \Foldl(f, f(x, y), ys)                  \\
      \Foldr(f, \Nil,         y) &= y                                       \\
      \Foldr(f, \Cons(x, xs), y) &= f(x, \Foldr(f, xs, y))
    \end{split}
  \end{equation*}
  \caption{A simple theory defining a $\List$ type and some associated
    operations, taken from~\cite{Johansson.Dixon.Bundy:conjecture-generation}.
    $\Zero$ and $\Succ$ are from a Peano encoding of the natural numbers.}
  \label{figure:list_theory}
\end{figure}

Given a logical theory, such as the theory of lists shown in
Figure~\ref{figure:list_theory}, we may want to find properties (e.g.
invariants) describing its behaviour. This could be for mathematical curiosity,
or due to the theory's importance in some domain. In particular, when dealing
with definitions in a software library, we may regard a set of such properties
as the code's \emph{specification}. Even those properties which only hold by
coincidence, rather than by design, may still be useful for \emph{optimising}
uses of this library, by rewriting expressions into equivalent forms which
require less time, memory, network usage, etc. For example, our theory of lists
turns out to satisfy the following invariant:

\begin{equation} \label{eq:mapreduce}
  \forall f. \forall xs. \forall ys.
    \Map(f, \Append(xs, ys)) = \Append(\Map(f, xs), \Map(f, ys))
\end{equation}

Whilst these two expressions produce equivalent results, the form on the right
can execute both calls to $\Map$ in parallel (in a pure functional language, at
least); a fact which is exploited by the ``map/reduce'' parallel programming
paradigm.

Such use cases require the ability to \emph{discover} true properties of some
particular definitions; this requires not only \emph{proving} their correctness,
but also \emph{posing} them as conjectures in the first place. This is a hard
problem in general, but presents opportunities for automation due to the
precise, symbolic nature of the domain. These processes can also feed into each
other, since proven properties can be used as lemmas for subsequent proofs. As
an example, proving Equation~\ref{eq:mapreduce} may require ``background
knowledge'', such as the property $\forall x. \Append(x, \Nil) = x$, which may
be tedious to produce by hand but possible to discover automatically.

\subsection{Automating Construction and Exploration of Theories}
\label{sec:te}

The task of automatically proving or disproving a \emph{given} conjecture has
been studied extensively in the field of Automated Theorem Proving (ATP). Less
attention has been paid to \emph{generating} such conjectures automatically,
either from a given set of definitions (the task we will focus on) or where the
definitions are also generated. This research has been pursued under the names
of Automated Theory Formation (ATF) and Mathematical Theory Exploration (MTE);
we will use the MTE terminology, and discuss their relationship further in
$\S$\ref{sec:related-work}.

A major challenge when generating conjectures is choosing how to narrow down the
resulting set to only those deemed ``interesting'', since this is an imprecise
term with many different interpretations. For example, all existing approaches
agree that simple tautologies are ``uninteresting'', but differ when it comes to
more complex statements.  Colton \etal{} surveyed tools for theory formation
and exploration, and their associated notions of ``interestingness'' for
concepts and conjectures~\cite{colton2000notion}. Six qualities were identified,
which are applied to conjectured properties as follows:

{\bf Empirical plausibility} checks whether a property holds across some
specific examples. This is especially useful for avoiding falsehoods, without
resorting to a deductive proof search.

{\bf Novelty} depends on whether the property, or one isomorphic or more
general, has already been seen.

{\bf Surprisingness} of a property is whether or not it is ``obvious'', for
example if it is an instance of a tautology.

{\bf Applicability} depends on the number of models in which a
property holds. High applicability makes a property more interesting, but so
does zero applicability (i.e. non-existence).

{\bf Comprehensibility} depends on the syntactic complexity of the property's
statement. Simpler statements are considered more interesting (which favours
tools adopting a small-to-large search order).

{\bf Utility} is the relevance or usefulness of a property to the user's
particular task. For example, if we want to find properties which are useful for
optimisation, like Equation~\ref{eq:mapreduce}, then utility would include
whether the property justifies some rewrite rule, the difference in resource
usage of the expressions involved, and how common those expressions are in real
usage.

Attempting to \emph{measure} such qualities is difficult, and having too many
measures complicates comparisons. System developers have employed a more
practical alternative in their evaluations, which is to perform
\emph{precision/recall} analysis against a \emph{ground-truth}. This requires
choosing a set of definitions (such as those in Figure~\ref{figure:list_theory})
and a set of properties (the ground truth) which represent the ``ideal'' result
of conjecture generation for those definitions (e.g. this might include
Equation~\ref{eq:mapreduce}). To analyse the quality of a tool's conjectures, we
run it on these chosen definitions and compare its output to the ground truth:

\begin{itemize}
\item \emph{Precision} is the proportion of a tool's output which appears in
  the ground truth. This penalises overly-liberal tools which output a large
  number of properties in the hope that some turn out to be ``good''.
\item \emph{Recall} is the proportion of the ground truth which appears in the
  tool's output. This penalises overly-conservative tools which generate very
  few properties as a way to avoid ``bad'' ones.
\end{itemize}

To score 100\% on precision and recall, all of the properties which appear in
the ground truth must have been conjectured, and nothing else. This gives us a
simple method to evaluate and compare tools without requiring a general solution
to the question of what is ``interesting''; although we must still decide what
to put in the ground truth set, and what to leave out, for each measured set of
definitions.

\begin{sloppypar}
  The precision and recall of three MTE tools are compared
  in~\cite{claessen2013automating}:
  IsaCoSy~\cite{Johansson.Dixon.Bundy:conjecture-generation},
  IsaScheme~\cite{Montano-Rivas.McCasland.Dixon.ea:2012} and
  HipSpec~\cite{Claessen_hipspec:automating}.\footnote{Unfortunately this
    comparison only reports prior results for each tool, rather than reproducing
    them. This makes the numbers less comparable, since each tool was tested
    with slightly different definitions (e.g. whether or not the exponential and
    maximum functions were included).} The definitions and ground truths for all
  three were taken from the standard library of the Isabelle proof assistant:
  since the library authors have gone to the effort of stating and proving these
  properties, and including them with every copy of their software, this is a good
  indication that they are interesting. Two sets of definitions were used: a
  theory of natural numbers; and a theory of lists (which we show in
  Figure~\ref{figure:list_theory} on page~\pageref{figure:list_theory}). With
  only two datapoints, each containing only a few definitions and properties (4
  functions and 12 properties, in the natural number example), it is difficult
  to draw general conclusions; yet despite these limitations we believe that
  this is a viable method to performing meaningful, objective analyses and
  comparisons for such tools. It is only with larger data sets that we can
  generalise the measured performance to different, especially \emph{novel},
  domains (which, after all, is the purpose of MTE tools). In addition, we might
  be concerned that such ``popular'' theories as the natural numbers may be
  unrepresentative of typical usage. For example, we would not expect the
  library functions in a typical verification problem to be as mathematically
  rich as number theory.
\end{sloppypar}

\subsection{Existing Tools}
\label{sec:existing-tools}

Since our benchmarking methodology builds on the precision/recall analysis
described in $\S$\ref{sec:te}, we decided to use two of the same tools, IsaCoSy
and QuickSpec (the conjecture generating component of HipSpec), as our
demonstration.

IsaCoSy (Isabelle Conjecture Synthesis) is written for the Isabelle proof
assistant, mostly in Standard
ML~\cite{Johansson.Dixon.Bundy:conjecture-generation}. It conjectures equations
by enumerating expressions involving a given set of (typed) constants and free
variables (a \emph{signature}). A constraint solver forbids certain
(sub)expressions from being synthesised, and these constraints are extended
whenever a new property is conjectured, to avoid generating any special-cases of
this property in the future. Conjectures are tested by looking for
counterexamples and, if none are found, sent to IsaPlanner which attempts to
prove them.

QuickSpec emerged from work on the QuickCheck software testing framework for the
Haskell programming language~\cite{claessen2011quickcheck}. As well as
conjecturing properties of Haskell definitions for HipSpec, it has been applied
to Isabelle definitions via the Hipster tool~\cite{Hipster}. Like IsaCoSy,
QuickSpec takes a signature and enumerates well-typed terms, collecting together
those of the same type into equivalence classes, assuming them to be
equal. QuickCheck is then used to find counterexamples to this assumption, which
it does by randomly instantiating the variables and comparing the resulting
closed expressions. Any equivalence class whose elements don't compare equal are
split up, and the process is repeated with new random values.

After 500 rounds of testing, any expressions still sharing the same equivalence
class are conjectured to be equal for all values of their variables. Rather than
using a constraint system to prevent redundancies (such as special-cases of
other properties), QuickSpec instead passes its output through a congruence
closure algorithm to achieve the same effect.

\section{Theory Exploration Benchmark}
\label{sec:proposal}

Our main contribution is a benchmarking methodology, shown in
Figure~\ref{figure:flow_chart} on page~\pageref{figure:flow_chart}, which both
generates a large definition/ground-truth corpus, and provides a scalable,
statistical approach to evaluating MTE tools using this corpus. We follow a
precision/recall approach similar to prior work, with the main difference being
the source of definitions and ground truths: we take existing problem sets
designed for automated theorem proving, and adapt their content for use in the
conjecture generation setting.

\subsection{Preparation}
\label{section:prep}

Automated theorem proving is an active area of research, with large problem sets
and regular competitions to prove as much as possible, as fast as
possible~\cite{pelletier2002development}. These problem sets are an opportunity
for MTE, as their definitions and theorems can be used as a corpus in the same
way that Isabelle libraries have been used in the past.

Some problem sets are more amenable for this purpose than others. The most
suitable are those meeting the following criteria:

\begin{itemize}
\item For each problem, there should be a clear distinction between the
  statement to be proved and the definitions involved, such that the two can be
  easily and meaningfully separated. This rules out problem sets like those of
  SMT-COMP~\cite{barrett2005smt}, where many problems involve uninterpreted
  functions, whose behaviour is \emph{implicit} in the logical structure of the
  theorem statement but not separately \emph{defined}.
\item Definitions should be translatable into a form suitable for the MTE tools
  under study. Our application in $\S$\ref{sec:application} requires Haskell and
  Isabelle translations, and also benefits from having definitions be strongly
  typed.
\item The problem set should be relevant to the desired domain. Our focus on
  functional programming requires higher-order functions and inductively defined
  datatypes, which rules out first-order languages/logics (such as
  TPTP~\cite{sutcliffe2009tptp}).
\item The problem set should ideally contain \emph{every} ``interesting''
  property involving its included definitions, since non-membership in the
  ground truth will be treated as being ``uninteresting''. More realistically,
  we should aim for each definition to have \emph{multiple} properties; rather
  than a mixture of unrelated problems, cherry-picked from different domains.
\item Larger problem set are preferable as they give more robust statistics, all
  else being equal (i.e. when this does not sacrifice quality).
\end{itemize}

Once such a problem set has been chosen, we must separate the theorem statements
from the definitions they reference. The latter will be used as input to the
MTE tools, whilst the former constitute the ground truth corpus of properties
against which to compare the output.

It is important to ensure that there are no duplicate definitions: we are only
concerned with the \emph{logical} content of the input, not the more arbitrary
aspects like the names of functions. For example, consider a problem set which
includes a statement of commutativity for a \texttt{plus} function, and of
associativity for an \texttt{add} function, where the definitions of
\texttt{plus} and \texttt{add} are $\alpha$-equivalent. We would expect an MTE
tool to either conjecture commutativity and associativity for \emph{both}
functions, or for \emph{neither} function, since they are logically equivalent.
Yet a na\"ive precision/recall analysis would treat commutativity of
\texttt{add} and associativity of \texttt{plus} as \emph{uninteresting}, since
they don't appear in the ground truth.

For this reason, duplicates should be removed, and any references to them
updated to use the remaining definition (e.g. chosen based on lexicographic
order). In the above example, the \texttt{plus} function would be removed, and
the commutativity statement updated to reference \texttt{add} instead.

\subsection{Sampling}
\label{section:sampling}

We could, in theory, send these de-duplicated definitions straight into an MTE
tool and use the entire set of properties (taken from the theorem statements) as
the ground truth for analysis. However, this would cause two problems:

\begin{itemize}
\item The result would be a single data point, which makes it difficult to
  infer performance \emph{in general}.
\item It is impractical to run existing MTE tools on ``large'' inputs,
  containing more than a few dozen definitions.
\end{itemize}

To solve both of these problems we instead \emph{sample} a subset of
definitions.\footnote{This could be done randomly, but for reproducibility we
  use a deterministic order (based on cryptographic hashes, to discourage
  ``cheating'').} Given a sample size, we choose a subset of that many definitions,
and provide only those as input to the tool. We generate a corresponding
ground truth by selecting those properties from the corpus which ``depend on''
(contain references to) \emph{only} the definitions in that sample. Transitive
dependencies aren't required (e.g. a property involving only a \texttt{times}
function would not depend on a \texttt{plus} function, even if \texttt{plus}
occurs in the definition of \texttt{times}).

Unfortunately, uniform sampling of definitions gives rise to a lottery: for a
given sample size, increasing the size of the corpus (which provides better
statistics) makes it less likely that a chosen sample will contain all of
a property's dependencies. The majority of such samples would hence have an empty
ground truth, and thus $0$ precision and undefined recall \emph{independent} of
the tool's output! This is clearly undesirable as an evaluation method.

Instead, we only allow a sample if it contains all dependencies of at least one
property. We could do this using rejection sampling, but it is more efficient to
pick a property, weighted in proportion to their number of dependencies
(ignoring those with more dependencies than our sample size). That property's
dependencies become our sample, padded up to the required size with uniform
choices from the remaining definitions.

The ground truth for such samples is guaranteed to contain at least one property
(the one we picked), and hence the precision and recall will depend meaningfully
on the tool's output.

\subsection{Evaluation}
\label{section:evaluation}

Given a sample of definitions and a corresponding ground truth, the actual
execution of the MTE tool proceeds as in prior work. We must translate the
chosen definitions into the required input format, then we time the execution
with a timeout (e.g. 5 minutes). We use wall-clock time since this is most
relevant to a user's experience, it has a straightforward interpretation and
tracking it does not require intrusive alterations to a tool's implementation.
The downside is that comparisons must take hardware performance into account,
which would not be the case if time were measured by some proxy like number of
expressions evaluated.

In our experiments we have found that memory usage is also an important part of
a tool's performance, but rather than complicating our analysis with an extra
dimension, we instead allow programs to use as much memory as they like, and
either get killed by the operating system or slow down so much from swapping
that they time out. This is in line with the expected usage of these tools:
either there is enough memory, or there isn't; implementations shouldn't be
penalised for making use of available resources.

To calculate precision and recall, the conjectures generated by each tool, as
well as the ground-truth properties, need to be parsed into a common format and
compared syntactically after normalising away ``irrelevant'' details. We
consider variable naming to be irrelevant, which can be normalised by numbering
free variables from left to right and using de Bruijn indices for bound
variables. We also consider the left/right order of equations to be irrelevant,
which we normalise by choosing whichever order results in the
lexicographically-smallest expression.

We specifically \emph{ignore} other logical relationships between syntactically
distinct statements, such as one equation being implied by another. Whilst
logically sound, second-guessing the ground truth in this way would harm other
aspects which influence interestingness (for example, a more general statement
is more widely applicable, but it might also be harder to comprehend).

This procedure gives us a single runtime, precision and recall value for each
sample. We propose two methods for analysing this data: \emph{summarising}
the performance of a single MTE tool and \emph{comparing} the performance of two
tools.

\subsection{Summarising}

Each sample is only explored once by each tool, so that we cover as many
independent samples as possible to better estimate how a tool's performance
generalises to unseen inputs. How we combine these data into an aggregate
summary depends on what we are interested in measuring. One general question we
might ask is how a tool's performance scales with respect to the input size (the
number of given definitions). This is straightforward to measure by varying the
sample size, but we need some way to combine the results from samples of the
same size.

We can summarise runtimes by choosing their median, as this is more robust than
the mean against long-running outliers, and hence represents performance for a
``typical'' input of that size. Unlike previous one-off evaluations, we can also
compute the \emph{spread} of our data, for example with the inter-quartile
range.

The precision and recall for multiple samples can be combined in two ways. We
can calculate their mean value, known as the \emph{average of the ratios}, to
find the expected precision and recall for runs of this size. Alternatively, we
can sum their numerators and denominators to get the \emph{ratio of the
  averages}: this normalises the output and ground truth cardinalities, allowing
precision and recall to be modelled as simple Bernoulli processes.

We prefer the latter, where each generated conjecture (wanted or not) is like a
coin toss whose bias is the precision; likewise, the ground truth properties
(found or not) are like coin tosses biased by the recall. Note that this simple
model does not explain how many conjectures the tool decides to output for each
run.

\subsection{Comparison}

To measure progress and encourage competition in the field, it is important to
compare the relative performance of different tools on the same task. Since the
aggregate statistics in the above summaries have obscured the details of
specific runs, any comparison based on them (such as comparing average recall)
would have very low statistical power. More direct comparisons can be made using
\emph{paired} tests, since for each individual sample we have measurements for
\emph{both} tools.

We propose that running times are compared using a Wilcoxon signed-rank
test~\cite{wilcoxon1945individual}, since it does not require that the data be
normally distributed. In $\S$\ref{sec:application} we follow a variant of the
Speedup-Test protocol of Touati, Worms and Briais~\cite{touati2013speedup}; the
differences are our use of the paired form of the Wilcoxon test, and the fact
that each run was performed on a different sample (to reduce sampling bias, at
the expense of accuracy). This paired difference test is also robust to the
censoring caused by timeouts: upper-bounding the run time gives an upper-bound
to the observable difference, which biases our results in a conservative way
(\emph{towards} the null hypothesis of indistinguishable performance).

For simplicity, we suggest comparing quality by pooling together results from
all samples where both tools finished successfully. Recall can be compared using
the standard McNemar test for paired samples~\cite{mcnemar1947note}: each
ground-truth property is categorised as either being found only by the first
tool, only by the second, by both or by neither. Those found by both or neither
are ignored, and the tool-specific numbers are tested for marginal homogeneity
(i.e. whether their differences give rise to similar totals).

We cannot use McNemar's test to compare precision, since each tool may generate
different conjectures. Instead, we can use Boschloo's form of Barnard's
test~\cite{lydersen2009recommended} to compare the proportions of wanted and
unwanted conjectures generated by each tool, assuming a binomial distribution
and conditioning on the number of generated conjectures (as if this were fixed
by design).

\begin{figure}
  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
                           minimum height=1cm,text centered, draw=black]

  \tikzstyle{io} = [trapezium, trapezium left angle=70,
                    trapezium right angle=110, minimum width=1cm,
                    minimum height=1cm, text centered, draw=black]

  \tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm,
                         text centered, draw=black]
  \tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm,
                          text centered, draw=black]

  \tikzstyle{arrow} = [thick,->,>=stealth]

  % Avoids too much padding in io shapes
  \tikzset{trapezium stretches=true}

  \centering
  \begin{tikzpicture}[node distance=1cm]

    % Preparation section

    \node (in)  [io]{Theorem Proving Benchmark};
    \node (sep) [process,       below=1cm of in         ]{Separate definitions from property statements};
    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};

    \node (thy)  [startstop, below=1cm of def]{Distinct Definitions};
    \node (thm)  [startstop, below=1cm of ref]{Corpus of Properties};

    \draw [arrow] (in)  -- (sep);
    \draw [arrow] (def) -- (ref);
    \draw [arrow] (def) -- (thy);
    \draw [arrow] (ref) -- (thm);

    % Arrows with labels
    \path[->]
        (sep) edge [arrow, sloped, above] node {definitions} (def)
        (sep) edge [arrow, sloped, above] node {properties}  (ref);

    % Sampling section

    \node (choose) [process, below=of thm   ]{Choose a property};
    \node (deps)   [process, below=of choose]{List property's dependencies};

    % Create dummy coordinate below deps, then use its y coordinate for pad
    \coordinate [below=of deps] (padDummy);
    \path let \p{dummy} = (padDummy),
              \p{in}    = (in)
              in coordinate (padPos) at (\x{in}, \y{dummy});
    \node (pad)  [process, at=(padPos)]{Pad list to form sample};

    \node (sthm) [startstop, below right=1.5cm and -1cm of pad]{Ground Truth};
    \node (sthy) [startstop, below  left=1.5cm and -1cm of pad]{Sampled Theory};

    % Calculate position of (size) using let, then define as normal
    \path let \p{pad}    = (pad),
              \p{choose} = (choose)
           in coordinate (sizePos) at (\x{pad},\y{choose});
    \node (size) [io, at=(sizePos)] {Sample size};

    % Evaluation section
    \path let \p{pad}  = (pad),
              \p{sthm} = (sthm)
           in coordinate (runPos) at (\x{pad}, \y{sthm});
    \node (run) [process, below=of runPos]{Run MTE tool};
    \node (pr)  [process, below=of run]{Analysis};

    \node (prec) [startstop, below=of pr  ]{Precision};
    \node (time) [startstop,  left=of prec]{Time taken};
    \node (rec)  [startstop, right=of prec]{Recall};

    \draw [arrow] (thy)    |- (pad);
    \draw [arrow] (thm)    -- (choose);
    \draw [arrow] (choose) -- (deps);
    \draw [arrow] (deps)   |- (pad);
    \draw [arrow] (size)   -- (choose);
    \draw [arrow] (size)   -- (pad);
    \draw [arrow] (pad)    -- (sthm);
    \draw [arrow] (pad)    -- (sthy);
    \draw [arrow] (sthy)   |- (run);
    \draw [arrow] (run)    -- (pr);
    \draw [arrow] (sthm)   |- (pr);
    \draw [arrow] (pr)     -- (prec);
    \draw [arrow] (pr)     -- (rec);
    \draw [arrow] (pr)     -- (time);

    % Awkward arrow
    \draw [arrow] (thm) -| ([shift={(7mm,-7mm)}]thm.east) |- (sthm);

    % Braces

    % Preparation
    \draw
      let \p{thm} = (thm.east),
          \p{in}  = (in.north),
          \p{rec} = (rec.south east)
       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
       (\x{rec}, \y{in}) -- (\x{rec}, \y{thm})
         node [black, midway, right, xshift=0.3cm]
              {Preparation $\S$\ref{section:prep}};

    % Sampling
    \draw
      let \p{thm} = (thm.east),
          \p{gt}  = (sthm.east),
          \p{rec} = (rec.south east)
       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
       (\x{rec}, \y{thm}) -- (\x{rec}, \y{gt})
         node [black, midway, right, xshift=0.3cm]
              {Sampling $\S$\ref{section:sampling}};

    % Evaluation
    \draw
      let \p{sthm} = (sthm.east),
          \p{rec}  = (rec.south east)
       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
       (\x{rec}, \y{sthm}) -- (\x{rec}, \y{rec})
         node [black, midway, right, xshift=0.3cm]
              {Evaluation $\S$\ref{section:evaluation}};

  \end{tikzpicture}
  \caption[]{High-level view of the benchmarking methodology described in
    $\S$\ref{sec:proposal}, showing
    \begin{tikzpicture}
      \node [rectangle, text centered, draw=black]{processes};
    \end{tikzpicture},
    \begin{tikzpicture}
      \node [rectangle, rounded corners, text centered, draw=black]{data};
    \end{tikzpicture} and
    \begin{tikzpicture}
      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
             text centered, draw=black]{inputs};
  \end{tikzpicture}}
  \label{figure:flow_chart}
\end{figure}

\section{Application}
\label{sec:application}

We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
benchmark as our ground truth corpus~\cite{claessen2015tip}.

To determine our benchmarking parameters we ran some initial tests on both tools
for a few samples sized between 1 and 100, for an hour each on our benchmarking
machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
of RAM. Most QuickSpec runs either finished within 200 seconds or not at all,
and sizes above 20 mostly timed out. IsaCoSy mostly finished within 300 seconds
on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
above this took thousands of seconds each, which we deemed infeasibly long.

Based on this we decided to benchmark sample sizes up to 20, since neither tool
seemed to perform well beyond that. The Speedup-Test protocol follows the
statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
so we pick 31 samples of each size in order to cross this threshold. This gave a
total of 1240 runs. To keep the benchmarking time down to a few days we chose a
timeout of 300 seconds, since that covered most of the successful QuickSpec and
IsaCoSy results we saw, and longer times gave rapidly diminishing
returns. During analysis, duplicate samples (caused by our requirement that
samples have a non-empty ground truth) were found and discarded, so only 14
samples of size 1 were used and 30 of size 2.

>>>>

%% FIXME: This is copypasta from benchmark paper
